A Standard for the Scholarly Citation of Archaeological Data: Supplementary Information

### 2017-08-20 analysis/paper/paper.Rmd
basically done with data analysis

We conducted a small pilot study to explore data sharing in archaeology. Our sample focused on _Journal of Archaeological Science_ articles on experiments with stone artefacts that were published during `r min(email_survey$'publication year', na.rm = TRUE)` and `r max(email_survey$'publication year', na.rm = TRUE)`. We focused on stone artefact experiments because the all research materials are generated by the researcher so there are not other stakeholders to consider when determining how to share the data: the researcher has complete control over the data. There are also no risks of damage to people or property in sharing these data because they do not include archaeological sites or human subjects. In contacting the authors, we requested the datasets for use in a graduate seminar with the aim of reproducing the published results and further exploring the data (this stage is still in progress). We promised to contact the authors if we discovered anything new, and not to share the data further without their permission. The data included with this paper is anonymised so the specific authors we contacted cannot be identified.

We sent out `r emails_out` emails to authors of these papers, and received `r emails_back`, resulting in a `r email_response_rate`% response rate. We recieved `r data_items_received` responses that included data files, giving an overall sharing rate of `r sharing_rate`%. Our small sample size limits the robustness of our analysis, but points to some observations that may be worth following up in a larger study. We found no clear relationship of data sharing and date of publication. We do not have any insights into why authors of older papers decline to share, but three of the six responses for papers published in 2014-2015 said they declined to share data because they intend to use these data in a future publication. Other noteworthy reasons provided by respondents for delining to share include the author about to defend their thesis or about to go on vacation. We interpret these to mean that the author does not have time to prepare their data for sharing, perhaps because the data are not already in a well-organised for ready for distribution. We found no effect on the author's student status at time of publication on data sharing behaviours, with roughly equal proportions of students and non-students sharing and not sharing. Similarly there are no clear patterns data sharing of among different journals. It is remarkable that the authors of the one PLOS One paper we contacted declined to share because that violates the policy of that journal at the time of publication of that paper (early 2014): "Publication is conditional upon the agreement of the authors to make freely available any materials and information described in their publication that may be reasonably requested by others for the purpose of academic, non-commercial research."

@@

file_format_combos_tally <- 
file_format_simple %>% 
  filter(!is.na(file_type)) %>% 
  group_by(n) %>% 
  filter(n_distinct(file_type) >= 2) %>% 
  slice(1) %>% # look at the file_format col only
  group_by(file_format) %>% 
  tally(sort = TRUE)
```

Our third investigation of archaeologists' data sharing behaviours is an analysis of all records (n = `r prettyNum(nrow(dc_arch), big.mark = ",")`, as of August 2017) of archaeological data files in online repositories tracked by DataCite (an organisation that provides persistent identifiers, or DOIs for research data). The majority of records come from the Archaeology Data Service (ADS) in the UK, followed by the Dutch Archiving and Network Services (DANS-EASY). Although the earliest data dates to `r first_date`, most of the data was collected after 1957. The was a steady increase in the amount of data deposited during 1990-2008, as personal computers become more affordable and data easier to digitize. There is a substantial increase in 2009 when the DANS-EASY repository saw in influx of records from Dutch CRM firms. This may be a follow-on effect of the introduction of Dutch archaeology legislation, the _Wet op de archeologische monumentenzorg_ (Wamz), in 2007 that formally obligated archaeologists in the Netherlands to archive their data via DANS. After 2009 the volume of deposits declined, with the exception of generic repositories such as Figshare, Zenodo and ResearchGate, which continued to grow.

Many of these records are grey literature such as unpublished reports, produced by commercial consultants or research projects. Of  the `r prettyNum(number_of_records_with_file_type, big.mark = ",")` (`r perc_of_records_with_file_type`%) records that include infomation about the file types in the deposit, `r perc_only_pdf`% (n = `r prettyNum(n_only_pdf, big.mark = ",")`) contain only PDFs. Structured data files such as spreadsheets, databases or shapefiles are found either alone or together with other file types (typically PDFs) in `r perc_structured_data_file`% (n = `r prettyNum(n_structured_data_file, big.mark = ",")`) records.

@@

rights_summary <- 
  rights %>% 
  mutate(the_text =  toupper(gsub(" |_|\\(|\\)", "-", Var1)),
         the_text = gsub("(\\-)\\1+", "\\1", the_text),
         the_text = gsub("-$", "", the_text)) %>% 
  mutate(cc = case_when(
    grepl("SEMANTICS/OPENACCESS", the_text) ~ "EU Open Access",
    grepl("SEMANTICS/RESTRICTEDACCESS", the_text) ~ "EU Restricted Access",
    grepl("SEMANTICS/CLOSEDACCESS", the_text) ~ "EU Closed Access",
    grepl("CC-BY-[^NC-ND]|CC-BY$|ATTRIBUTION-LICENSE|ATTRIBUTION-[^NON]", the_text) ~ "CC-BY",
    grepl("CC-BY-NC-[^ND]|BY-NC-[^ND]|CC-NC$|ATTRIBUTION-NON-COMMERCIAL-", the_text) ~ "CC-BY-NC",
    grepl("CC-BY-NC-SA-[^ND]|CC-NC-SA$|ATTRIBUTION-NON-COMMERCIAL-SHARE", the_text) ~ "CC-BY-NC-SA",
    grepl("CC-BY-NC-ND|BY-NC-ND|ATTRIBUTION-NONCOMMERCIAL-NO-DERIVATIVE|ATTRIBUTION-NONCOMMERCIAL-NODERIVATIVES|ATTRIBUTION-NONCOMMERCIAL-NODERIVS|NAMENSNENNUNG-NICHTKOMMERZIELL-KEINEBEARBEITUNG|NAMENSNENNUNG-KEINE-KOMMERZIELLE-NUTZUNG", the_text) ~ "CC-BY-NC-ND",
    grepl("ADS|ARCHAEOLOGYDATASERVICE", the_text) ~ "ADS",
    grepl("CC0|CC-0|ZERO", the_text) ~ "CC-0"
  )) %>% 
  select(cc, n) %>% 
  filter(!is.na(cc)) %>% 
  group_by(cc) %>% 
  tally(sort = TRUE)

@@

# show plots
datacentre_plot
allocator_plot
creator_plot
language_plot
contributor_plot
rights_plot
publisher_plot

>>>>

Our sample focused on _Journal of Archaeological Science_ articles on experiments with stone artefacts that were published during `r min(email_survey$'publication year', na.rm = TRUE)` and `r max(email_survey$'publication year', na.rm = TRUE)`. We focused on stone artefact experiments because the all research materials in these projects are generated by the researcher, so there are no other stakeholders to consider when determining how to share the data: the researcher has complete control over the data. There are also no risks of damage to people or property in sharing these data because they do not include archaeological sites or human subjects. In contacting the authors, we requested the datasets for use in a graduate seminar with the aim of reproducing the published results and further exploring the data (this stage is still in progress). We promised to contact the authors if we discovered anything new, and not to share the data further without their permission. The data included with this paper is anonymised so the specific authors we contacted are not identifiable.

We sent out emails to the first authors of `r emails_out` stone artefact experiment papers, and recieved `r emails_back` replies, resulting in a `r email_response_rate`% response rate. We recieved `r data_items_received` responses that included data files, giving an overall sharing rate of `r sharing_rate`%. Our small sample size limits the robustness of our analysis, but points to some observations that may be worth following up in a larger study. We found no clear relationship of data sharing and date of publication. We do not have any insights into why authors of older papers decline to share, but three of the six responses for papers published in 2014-2015 said they declined to share data because they intend to use these data in a future publication. Other noteworthy reasons provided by respondents for declining to share include the author about to defend their thesis or about to go on vacation. We interpret these to mean that the author does not have time to prepare their data for sharing, perhaps because the data are not already in a well-organised for ready for distribution. We found no effect on the author's student status at time of publication on data sharing behaviours, with roughly equal proportions of students and non-students sharing and not sharing. Similarly there are no clear patterns data sharing of among different journals. It is remarkable that the authors of the one PLOS One paper we contacted declined to share because that refusal violates the policy of that journal at the time of publication of that paper (early 2014): "Publication is conditional upon the agreement of the authors to make freely available any materials and information described in their publication that may be reasonably requested by others for the purpose of academic, non-commercial research."

@@

file_format_combos_tally <- 
file_format_simple %>% 
  filter(!is.na(file_type)) %>% 
  group_by(n) %>% 
  filter(n_distinct(file_type) >= 2) %>% 
  slice(1) %>% # look at the file_format col only
  group_by(file_format) %>% 
  tally(sort = TRUE)

# which group has the highest *proportion* of structured data files?
proportion_structured_data_files <- 
dc_arch %>% 
    mutate(strutured_data = ifelse(grepl("access|mdb|dbf|dbase|accdb|dbase|arcgis|shape|shp|mapinfo|mid|mif|surfer|xls|excel|csv|spreadsheet|spreadsheet", tolower(format)), "has_structured_data", "no")) %>% 
  group_by(datacentre, strutured_data) %>% 
  tally(sort = TRUE) %>% 
  spread(strutured_data, n) %>% 
  mutate(perc_structured = has_structured_data / (has_structured_data + no) * 100) %>% 
  arrange(desc(perc_structured))

perc_struc_in_ads <- 
  proportion_structured_data_files %>% 
  filter(grepl("ADS", datacentre)) %>% 
  pull(perc_structured) %>% 
  round(., 2)

n_struc_in_ads <- 
  proportion_structured_data_files %>% 
  filter(grepl("ADS", datacentre)) %>% 
  pull(has_structured_data)

perc_struc_in_dans <- 
  proportion_structured_data_files %>% 
  filter(grepl("DANS", datacentre)) %>% 
  pull(perc_structured) %>% 
  round(., 0)

n_struc_in_dans <- 
  proportion_structured_data_files %>% 
  filter(grepl("DANS", datacentre)) %>% 
  pull(has_structured_data)

```{r rights, eval = TRUE}
rights_summary <- 
  rights %>% 
  mutate(the_text =  toupper(gsub(" |_|\\(|\\)", "-", Var1)),
         the_text = gsub("(\\-)\\1+", "\\1", the_text),
         the_text = gsub("-$", "", the_text)) %>% 
  mutate(cc = case_when(
    grepl("SEMANTICS/OPENACCESS", the_text) ~ "EU Open Access",
    grepl("SEMANTICS/RESTRICTEDACCESS", the_text) ~ "EU Restricted Access",
    grepl("SEMANTICS/CLOSEDACCESS", the_text) ~ "EU Closed Access",
    grepl("CC-BY-[^NC-ND]|CC-BY$|ATTRIBUTION-LICENSE|ATTRIBUTION-[^NON]", the_text) ~ "CC-BY",
    grepl("CC-BY-NC-[^ND]|BY-NC-[^ND]|CC-NC$|ATTRIBUTION-NON-COMMERCIAL-", the_text) ~ "CC-BY-NC",
    grepl("CC-BY-NC-SA-[^ND]|CC-NC-SA$|ATTRIBUTION-NON-COMMERCIAL-SHARE", the_text) ~ "CC-BY-NC-SA",
    grepl("CC-BY-NC-ND|BY-NC-ND|ATTRIBUTION-NONCOMMERCIAL-NO-DERIVATIVE|ATTRIBUTION-NONCOMMERCIAL-NODERIVATIVES|ATTRIBUTION-NONCOMMERCIAL-NODERIVS|NAMENSNENNUNG-NICHTKOMMERZIELL-KEINEBEARBEITUNG|NAMENSNENNUNG-KEINE-KOMMERZIELLE-NUTZUNG", the_text) ~ "CC-BY-NC-ND",
    grepl("ADS|ARCHAEOLOGYDATASERVICE", the_text) ~ "ADS",
    grepl("CC0|CC-0|ZERO", the_text) ~ "CC-0"
  )) %>% 
  select(cc, n) %>% 
  filter(!is.na(cc)) %>% 
  group_by(cc) %>% 
  tally(sort = TRUE) %>% 
  mutate(perc = round(nn / sum(nn) * 100, 0))

# what proportion is ADS or others?
rights_ads_perc <- 
rights_summary %>% 
  filter(cc == "ADS") %>% 
  pull(perc)

rights_ads_n <- 
rights_summary %>% 
  filter(cc == "ADS") %>% 
  pull(nn)

rights_eu_perc <- 
rights_summary %>% 
  filter(cc == "EU Open Access") %>% 
  pull(perc)

rights_eu_n <- 
rights_summary %>% 
  filter(cc == "EU Open Access") %>% 
  pull(nn)

Our third investigation of archaeologists' data sharing behaviours is an analysis of all records (n = `r prettyNum(nrow(dc_arch), big.mark = ",")`, as of August 2017) of archaeological data files in online repositories tracked by DataCite (an organisation that provides persistent identifiers, or DOIs for research data). The majority of records come from the Archaeology Data Service (ADS) in the UK, followed by the Dutch Archiving and Network Services (DANS-EASY). Although the earliest data dates to `r first_date`, there was a steady increase in the amount of data deposited during 1990-2008, as personal computers become more affordable and data easier to digitize. There is a substantial increase in 2009 when the DANS-EASY repository saw in influx of records from Dutch CRM firms. This relates to the introduction of Dutch archaeology legislation, the _Wet op de archeologische monumentenzorg_ (Wamz), in 2007 that formally obligated archaeologists in the Netherlands to archive their data via DANS within two years of completion of their projects (Keers et al. 2011). After 2009 the volume of deposits declined in many repositories, with the exception of generic repositories such as Figshare, Zenodo and ResearchGate, which continued to grow.

Of the `r prettyNum(sum(rights_summary$nn), big.mark = ",")` records with copyright information, the majority are released under the ADS conditions of use (n = `r prettyNum(rights_ads_n, big.mark = ",")`, `r rights_ads_perc`%), which is broadly similar to the CC-BY-NC (Moore and Richards 2015). The next most frequently used condition is the EU Open Access licence (n = `r prettyNum(rights_eu_n, big.mark = ",")`, `r rights_eu_perc`%), similar to the CC-BY licence. Creative Commons licenses such as CC-BY, CC-0 and others are used by less than 3% of records.

Many of the records are grey literature such as unpublished reports, produced by commercial consultants or research projects. Of the `r prettyNum(number_of_records_with_file_type, big.mark = ",")` (`r perc_of_records_with_file_type`%) records that include information about the file types in the deposit, `r perc_only_pdf`% (n = `r prettyNum(n_only_pdf, big.mark = ",")`) contain only PDFs. Structured data files such as spreadsheets, databases or shapefiles are found either alone or together with other file types (typically PDFs) in `r perc_structured_data_file`% (n = `r prettyNum(n_structured_data_file, big.mark = ",")`) records. The highest proportion of records with structured data files are found in the DANS-EASY repository (n = `r prettyNum(n_struc_in_dans, big.mark = ",")`, `r perc_struc_in_dans`%), followed by the ADS with `r perc_struc_in_ads`% (n = `r n_struc_in_ads`). Note that we do not have any file format information for all the records from some smaller repositories in this sample, so it is possible that the true proportion of structured files in these records is higher.

@@

@@

# show plots
datacentre_plot
# allocator_plot
# creator_plot
language_plot
# contributor_plot
rights_plot
# publisher_plot
